---
title: "Data Preprocessing"
author: "Corrie"
date: "February 23, 2019"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

## Preprocessing The Data
We will be using the data provided by the World Happiness Report. The report from 2018 can be found [here](https://s3.amazonaws.com/happiness-report/2018/CH2-WHR-lr.pdf). 

We first read in the data and change the column names to names easier to work with.
```{r, message=F, warning=F}
library(tidyverse)

d <- read_csv("../data/WHR2018Chapter2OnlineData.csv") %>%
  rename(happiness =`Life Ladder`, 
         log_gdp =`Log GDP per capita`, 
         social = `Social support`, 
         health = `Healthy life expectancy at birth`, 
         freedom = `Freedom to make life choices`, 
         generosity = Generosity, 
         corruption = `Perceptions of corruption`, 
         pos_affect = `Positive affect`, 
         neg_affect = `Negative affect`, 
         gov_confidence = `Confidence in national government`, 
         democracy = `Democratic Quality`, 
         delivery = `Delivery Quality`, 
         sd_happiness = `Standard deviation of ladder by country-year`, 
         sd_mean_happiness = `Standard deviation/Mean of ladder by country-year`,
         gini = `GINI index (World Bank estimate)`, 
         gini_avg = `GINI index (World Bank estimate), average 2000-15`, 
         gini_household = `gini of household income reported in Gallup, by wp5-year`)
```
The supporting data also includes a region indicator for the different countries. We can roughly sort these region indicators to their continents which makes it easier for plotting later. Note that Australia and New Zealand are in the same region as the US and Canada and thus landed on the continent Americas.
```{r, message=F, warning=F}
missing_regions <- c("Comoros" = "Sub-Saharan Africa",
  "Cuba" = "Latin America and Caribbean",
  "Djibouti" = "Middle East and North Africa",
  "Guyana" = "Latin America and Caribbean",
  "Oman" = "Middle East and North Africa",
  "Somaliland region" = "Middle East and North Africa",
  "Suriname" = "Latin America and Caribbean",
  "Swaziland" = "Sub-Saharan Africa")

d2 <- read_csv("../data/WHR2018Chapter2OnlineData_supporting_factors.csv") %>%
  select(country, region = `Region indicator`) 



d <- d %>%
  left_join( d2, by="country") %>%
  mutate(region = ifelse(is.na(region), country, region)) %>%
  mutate(region = str_replace_all(region, missing_regions)) %>%
  mutate(continent = fct_collapse(region,
                                  America=c("North America and ANZ", "Latin America and Caribbean"),
                                  Africa=c("Middle East and North Africa", "Sub-Saharan Africa" ),
                                  Asia=c("South Asia", "Southeast Asia", "East Asia", 
                                         "Commonwealth of Independent States"),
                                  Europe=c("Central and Eastern Europe", "Western Europe")
                                  )) 
```

The World Happiness Report only provides the Log of the GDP and by the time of publication, the GDP of 2017 wasn't available yet and extrapolated in the report. We can use the updated version of the GDP as provided by the [World Bank](https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD). We use the GDP per capita in purchasing power parity
(PPP) at constant 2011 international dollar prices.
```{r, message=F, warning=F}
country_spell_list <- c("Egypt, Arab Rep."="Egypt",
   "Russian Federation"="Russia",
   "Korea, Rep."="South Korea",
   "Yemen, Rep."="Yemen",
   "Slovak Republic"="Slovakia",
   "Iran, Islamic Rep."="Iran",
   "Congo, Rep."="Congo (Brazzaville)",
   "Congo, Dem. Rep."="Congo (Kinshasa)",
   "Hong Kong SAR, China"="Hong Kong S.A.R. of China",
   "Cote d'Ivoire"="Ivory Coast",
   "Kyrgyz Republic"="Kyrgyzstan",
   "Macedonia, FYR"="Macedonia",
   "Lao PDR"="Laos",
   "Venezuela, RB"="Venezuela",
   "Syrian Arab Republic"="Syria")

gdp <- read_csv("../data/worldbank_gdp.csv",
                skip=3) %>%
  rename(country=`Country Name`, country_code=`Country Code`) %>%
  select(country, country_code, `1960`:`2017`) %>%
  gather(key="year", value="gdp", `1960`:`2017`) %>%
  mutate(log_gdp = log(gdp),
         year=as.numeric(year)) %>%
  mutate(country = str_replace_all(country, country_spell_list))



d <- d %>%
  select(-log_gdp) %>% 
  left_join( gdp, by=c("country", "year")) %>%
  select(country, country_code, region, continent,
         year, happiness, gdp, log_gdp, social:gini_household)
```
Some countries are not in the GDP data, so we manually add the country code:
```{r}
missing_country_codes <- c("Swaziland"="SWZ",
  "Palestinian Territories"="PSE",
  "North Cyprus"="CYP",
  "Somaliland"="SOM",
  "Taiwan Province of China"="TWN")

d <- d %>% 
 mutate(country_code = ifelse(is.na(country_code), country, country_code)) %>%
  mutate(country_code = str_replace_all(country_code, missing_country_codes))
  
```


## Adding Further Variables
We can easily add further data.

### Add Population Data
 For example, the World Bank also provides yearly [population data](https://data.worldbank.org/indicator/sp.pop.totl)
```{r, warning=F, message=F}
pop <- read_csv("../data/worldbank_population.csv",
                skip=3) %>%
  select(country=`Country Name`, `1960`:`2017`) %>%
  mutate(country = str_replace_all(country, country_spell_list)) %>%
  gather(key="year", value="pop", `1960`:`2017`) %>%
  mutate(year = as.numeric(year),
         log_pop = log(pop))
  
d <- d %>%
  left_join(pop, by=c("country", "year"))
```

### Add Temperature and Rain Data
We can add [Climate Data](https://datacatalog.worldbank.org/dataset/climate-change-knowledge-portal-historical-data):

```{r}
temp <- read_csv("../data/temperature_data.csv") %>%
  select(country_code=ISO_3DIGIT, avg_temp=Annual_temp)

rain <- read_csv("../data/precipitation_data.csv") %>%
  select(country_code=ISO_3DIGIT, annual_precip=Annual_precip)

d <- d %>%
  left_join(temp, by="country_code") 
  
d <- d %>%
  left_join(rain, by="country_code") 
```


## Save Data Frame
```{r}
write_csv(d, path="../data/preprocessed_data.csv")
```

